\documentclass[11pt,letterpaper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}

\pagestyle{empty}

\begin{document}

\today

\vspace{1cm}

Dear Editor,

\vspace{0.5cm}

We submit for your consideration the manuscript entitled ``\textbf{AR(2) Eigenvalue Modulus as a Measure of Temporal Persistence in Gene Expression: Circadian Hierarchy Emerges from Two Coefficients}'' for publication in \textit{PLOS Computational Biology}.

\vspace{0.3cm}

\textbf{Summary.} We demonstrate that the eigenvalue modulus of a second-order autoregressive model---a standard tool from econometrics and control theory---measures temporal persistence in gene expression and blindly recovers the known circadian hierarchy (clock $>$ target $>$ other) across 4 species, 12 mouse tissues, and 36 tissue- or condition-level time series from 10 independent GEO datasets.

\vspace{0.3cm}

\textbf{Significance.} No single metric has previously captured circadian hierarchy from expression data alone. Existing tools (JTK\_CYCLE, cosinor, RAIN) detect \textit{rhythmicity}; our eigenvalue modulus measures \textit{persistence}---a complementary dynamical property. We go beyond standard validation with five lines of evidence: (1) a twelve-analysis robustness suite including BMAL1-knockout causal validation; (2) cross-referencing against 59 literature-validated circadian genes (98.3\% recovery); (3) a Popper-faithful falsification test showing $\sim$180-fold enrichment of BMAL1 coupling over housekeeping controls; (4) three automated bias audits confirming the hierarchy depends on temporal order, not confounds; and (5) extension to non-circadian biology (dendritic cell immune response) validating the principle beyond circadian rhythms. The eigenvalue modulus is independent of mRNA half-life ($\rho = 0.012$ across 22,989 genes), confirming it measures a genuinely novel property.

\vspace{0.3cm}

\textbf{Reproducibility.} All 10 source datasets are publicly available from NCBI GEO. Complete AR(2) results for all genes are provided as supplementary files. An interactive web application (\url{https://par2-discovery-engine.replit.app}) provides access to all results and allows users to upload their own data.

\vspace{0.3cm}

\textbf{Limitations acknowledged.} We are transparent about the method's boundaries: baboon replication is partial (7/8 directionally correct, 4/8 bootstrap-significant, consistent with the 12-timepoint limitation); the fine-grained nine-category hierarchy beyond the clock--other gap is suggestive rather than definitive; and we recommend $N \geq 16$ timepoints for robust results. Twelve limitations are discussed in the manuscript.

\vspace{0.3cm}

This manuscript has not been submitted elsewhere and all work is original.

\vspace{0.5cm}

Sincerely,

\vspace{0.5cm}

Michael Whiteside \\
Independent Researcher, United Kingdom \\
mickwh@msn.com \\
ORCID: 0009-0000-0643-5791

\end{document}
