\documentclass[11pt,letterpaper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}

\pagestyle{empty}

\begin{document}

\today

\vspace{1cm}

Dear Editor,

\vspace{0.5cm}

We submit for your consideration the manuscript entitled ``\textbf{A 12-Tissue Atlas of Circadian Gene Coupling: Genome-Wide BMAL1 Predictor Screening Reveals Universal and Tissue-Specific Clock Connections}'' for publication in \textit{Molecular Systems Biology}.

\vspace{0.3cm}

\textbf{Summary.} We present a systematic atlas of BMAL1 (Arntl) statistical coupling across 12 mouse tissues and genome-wide in mouse liver. Using AR(2)+exogenous predictor modeling, we identify 63 BMAL1-coupled genes genome-wide (180-fold enrichment over random predictors, $p < 10^{-10}$) and 85 significant coupling events across 33 genes in the 12-tissue atlas. Wee1 is coupled in 10/12 tissues (the most broadly conserved circadian output), and Nampt in 8/12 tissues.

\vspace{0.3cm}

\textbf{Significance.} This work provides the most comprehensive statistical coupling atlas of the core circadian transcription factor BMAL1 across mammalian tissues. The decisive falsification test (180-fold enrichment over housekeeping and random gene predictors) rules out statistical artifact. Seven predictions are independently confirmed by published wet-lab experiments. Ten novel predictions---including circadian coupling of the stem cell marker LGR5---await experimental validation.

\vspace{0.3cm}

\textbf{Clinical relevance.} The near-universal coupling of Wee1 across tissues predicts that Wee1 inhibitor efficacy (e.g., adavosertib, in active clinical trials) should be strongly time-of-day dependent. Tissue-specific coupling programs suggest that chronotherapy strategies must be tailored to target tissue.

\vspace{0.3cm}

This manuscript has not been submitted elsewhere and all work is original.

\vspace{0.5cm}

Sincerely,

\vspace{0.5cm}

Michael Whiteside \\
Independent Researcher, United Kingdom \\
mickwh@msn.com \\
ORCID: 0009-0000-0643-5791

\end{document}
